---
title: "Preparing Census data for SMS"
author: 'Author: Néstor'
subtitle: Data cleaning and transformation
output: 
  github_document:
    toc: true
    toc_depth: 2
---

`r load(file = "C:/1.population.modelling/population.modelling/results/data/preparing.census.data.for.sms.RData")`
`r library(knitr)`
`r library(ggplot2)`

# 1. Introduction

This document shows the steps taken for preparing the census data for SMS. Census data related to `age-ranges, sex, school, and work` allows the estimations DWW pollutants (production and location) depending on population characteristics. 


## 1.1 Variables selected


After analyzing **`r paste(ncol(hpconsfull))`** census variables that exist at the block level, **`r paste(ncol(hpcons.orig))`** were required in the phase of `cleaning and transformation` to execute SMS and to provide input data for the DWW ABM. The selected variables will be used to simulate the dynamics of the `mobility`, `biological`, and `social` behaviors as `ABM's submodels`.

```{r echo=FALSE}
colnames(hpcons.orig)
```

## 1.2 `NA` Values

*NA* cells exits as a mean of privacy protection. The raw data by block demonstrate presence of *NA* values equal to **`r paste((sum(is.na(hpcons.orig))))`** cells. A preview of data missing due to privacy looks like the following:


```{r echo=FALSE}
hpcons.orig[,7:15]
```
<br>

## 1.3 `NA` Confidentiality criteria

The main confidentiality criteria defined by the *Ley del Sistema Nacional de Información Estadística y Geográfica (LSNIEG)* are the following (INEGI,2020):

    • At the municipality or territorial demarcation level, locality and AGEB, any indicator with less than three units appears with an asterisk with the exception of the variables Total population (POBTOT), Total dwellings (VIVTOT) and Total inhabited dwellings (TVIVHAB).

    • At the block level, the previous criterion also applies; additionally, for those with one or two inhabited dwellings, only presents information in the variables Total population (POBTOT) and Total dwellings (VIVTOT), in the rest of the indicators appear asterisks.

The above criteria is considered to fill NA cells. Note that in this document `*` are replace by `NA` cells.


# 2 *R script* rules

The prediction of DWW pollutants depends on the quantification of population totals and their characteristics. `NA` cells represent missing population that leads to an underestimation of pollutants, for that reason `NA` cells are replaced with values following the rules in this section. Once values are filled, it is possible to fulfill the requirements to execute SMS for a further estimation of DWW pollutants.


## 2.1 Functions to fill `NA` cells at each categorical variable.

As a general overview, `NA` cells are filled with a number of population that matches the totals:

Targeted totals  | Description
------------- | -------------
By column   | Representing the totals of the categorical values.
By row  | Making sure that the totals a block are not overestimated.
By gender  | Consideration in the desagregation of the totals by gender (and their sum).

The variables selected represent the following categories:

  * `Population totals` 
  * `Age-ranges and sex`
  * `Assistance or not to school`
  * `Actively economically or not (work)`

Each category requires its own script for data cleaning and transformation. This document explains the categorical variable `Age-ranges and sex`. The functions for the rest of the categorical variables are adjusted based on the similar logic. If there is interest, it is also possible to explore the code for a better understanding of the particularities of each script by targeted category.

The following script take a glance of the core rules for cleaning `Age-ranges and sex` (code description provided after the script):

    To interpret the code it is relevant to note the following terminology:

Notation in the script  | Description
------------- | -------------
`smcol_, smrow_`  | Built functions to sum rows-columns across the categorical variable.
`_mf, _m, _f`  | Endings that refer at each section of a function by gender.
`.mf, .m, .f`  | If contained, it refers at each variable by its gender.
`vec.mf.n, vec.m.n, vec.f.n`  | Vectors variables that saves the new `.n` values.
`t.tot`  | Totals of the categorical value (columns).
`hpcons`  | Original data table with all categorical variables.
`df.c$col.f`  | Ending that refers to a cleaned and/or calculated dataframe and variable.


```{r highlight=TRUE, eval=FALSE}
##########################################
#RULES FOR TOTALS BY SUM OF MALE AND FELAME (MF)
##########################################

#Conditionals for totals MF: Filling missing values
  while (t.tot.mf > smcol_mf(i)){#while vales are not == real total population
    for (i in 1:nrow(hpcons)){#For each cell in the target variable column
      if (is.na(vec.mf[i])#Targeting the missing fields to fill
          &(t.tot.mf > smcol_mf(i))#Keep sum col values under real total
          &(POBTOT.c$POBTOT.f[i] > smrow_mf(i))#keep row sum variable under POB totals
          &(3 > vec.mf.n[i])#Only values less than 3 are confidential
      ) {vec.mf.n[i] <- vec.mf.n[i]+1}#Add individual
    }
  }

##########################################
#RULES FOR TOTALS BY MALE (M)
##########################################

#Conditionals for totals M: Filling missing values
#First make sure to cover all female conditions
  while (t.tot.m > smcol_m(i)){#while vales are not = real total population
    for (i in 1:nrow(hpcons)){#For each cell in the target variable column
      if (is.na(vec.m[i])#Targeting the missing fields to fill
          & !is.na(vec.f[i])#For female precondition for male cells
          & (POBTOT.c$POBMAS.f[i] > smrow_m(i))#Keep sum col values under real total
          #First fill all values where total and females are defined
          & (((vec.mf + vec.mf.n)[i] - vec.f[i]) != vec.m.n[i])
      ){while (((vec.mf + vec.mf.n)[i] - vec.f[i]) != vec.m.n[i]){
        vec.m.n[i] <- vec.m.n[i]+1
        }
      }
    }
    for (i in 1:nrow(hpcons)){
      if(is.na(vec.m[i])#Targeting the missing fields to fill
         & is.na(vec.f[i])#2nd possibility with female values
         & (t.tot.m > smcol_m(i))#Keep sum col values under real total
         & (POBTOT.c$POBMAS.f[i] > smrow_m(i))#Keep sum row values under real total
         & (vec.m.n[i] != vec.df.mf.n[i,c(1)])#must be different that the MF total of the cell
         & (3 > vec.m.n[i])#Only values less than 3 are confidential
      ){#keep by block values under POB totals
        vec.m.n[i] <- vec.m.n[i]+1#Add individual
      }
    }
  }

##########################################
#RULES FOR TOTALS BY FEMALE (F)
##########################################

#Conditionals for totals F: Filling missing values
  while (t.tot.f > smcol_f(i)){#while vales are not = real total population
    for (i in 1:nrow(hpcons)){#For each cell in the target variable column
      if (is.na(vec.f[i])#Targeting the missing fields to fill
          & (t.tot.f > smcol_f(i))#Keep sum col values under real total
          & (POBTOT.c$POBFEM.f[i] > smrow_f(i))#Keep sum row values under real total
          #Special condition: verify that male[i] keeps under total MF[i]
          & ( (vec.mf+vec.mf.n)[i] != ((vec.m+vec.m.n)[i] + vec.f.n[i]))
          #Special condition: Calculate difference of new totals
          & (( (vec.mf+vec.mf.n)[i] - (vec.m+vec.m.n)[i]) != 0)
          & (3 > vec.m.n[i]))#Only values less than 3 are confidential
      {vec.f.n[i] <- vec.f.n[i]+1#Add individual
      }
    }
  }
```

As can be seen in the script headers, the rules are separated in three sections, 1. Sum of male and female, 2. Male, 3. Female. All categorical variables are disaggregated by gender, and the structure of their functions follows the same structure.

It can be noted that for the three sections, the script of `Male + Female` is repeated for `Male` and `Female` except that some additions are required depending on the gender column. For that reason the explanation of the `Male + Female` code in the following section is enough to understand the rest of the code.

It is relevant to mention that the order to execute the code is relevant. They exist differences on the `Male` and `Female` script that are discussed in the following section. For the rest, the design of the script is self-explained.

## 2.2 Columns(`smcol_`) Targets the consistency in the totals by categorical variable.

This section describes the function `smcol_`. The function: `while(t.tot.mf > smcol_mf(i))` is the initial condition in each section of the code. The function makes sure that `NA` vales are filled until matching the totals of the categorical variable.

### Tables: `t.tot` & `variable table`

Each categorical variable consist of three columns separated by their count of totals (male + female), count of totals by male, and by female, e.g;`P_18A25`, `P_18A25_M`, `P_18A25_F`. The total table (`t.tot`) consist of one row, and number of columns equal to the categorical variables. Example of the data tables:

```{r echo=FALSE}
kable(hptotcons[,37:39], caption = "Section of `t.tot` with the totals by a categorical variable (columns):")
```

```{r echo=FALSE}
kable(hpcons[1:6,37:39], caption = "Section of the `variable table` with NA values for a categorical variable (block data):")
```
### The `while` conditionals

As the census data provides the total population for each categorical variable in a locality, it is possible to add values into `NA` cells until matching the total criteria in the locality of study. For that purpose a `while` conditional is used for each categorical variable and their three columns:

```{r highlight=TRUE, eval=FALSE}

Totals:   while (t.tot > smcol(i)), vec.n[i] <- vec.n[i]+1
Male:     while (t.tot.m > smcol_m(i)), vec.m.n[i] <- vec.m.n[i]+1
Female:   while (t.tot.f > smcol_f(i)), vec.f.n[i] <- vec.f.n[i]+1
```

Where:  | Description
------------- | -------------
`t.tot`  | Population totals of the categorical variable.
`smcol(i)`  | Sum of column values of the variable in the cell iteration `i`.
`vec.n`  | New vector value in the cell `i` that replace the `NA`.
`vec.n[i]+1`  | Adds a unit in the new vector cell.
    
For each iteration (i) in a cell of the column, the script targets an `NA` cell, adds `1` and goes to the next `NA` cell to repeat the process until the totals of the categorical variables are not longer bigger than the sum of the column by block. As a result, the population for targeted categorical variable fills the missing population at the blocks.

### The `if` conditionals

There are some conditionals that must be true in each iteration of the cells to define if a value can be added at the cell `i` during the `while` loop, which are the following:

```{r highlight=TRUE, eval=FALSE}
if(is.na(vec.mf[i])#Targeting the missing fields to fill
&(t.tot.mf > smcol_mf(i))#Keep sum col values under real total
&(POBTOT.c$POBTOT.f[i] > smrow_mf(i))#keep row sum variable under POB totals
&(3 > vec.mf.n[i])#Only values less than 3 are confidential
```

The `is.na` function make sures that only `NA` cells are selected. The `(t.tot.mf > smcol_mf(i))` function is required to make sure that during an specific iteration `i` such condition have been fulfilled or not. The function `(3 > vec.mf.n[i])` express the confidentiality criteria. Finally the function `(POBTOT.c$POBTOT.f[i] > smrow_mf(i))` is explained in the following section.



## 2.3 Rows (`smrow_`): Consistency in the totals by block.

The totals by row allows to control a gradual addition of individuals in `NA` cells under the threshold of the total population in each block for every categorical variable. 

### Tables: `POBTOT` & `variable table`

In each block (rows), totals by gender are provided for locality, which can be appreciated in the following table:

```{r echo=FALSE}
knitr::kable(
  list((hpcons[1:7,1:3]), (hpcons[1:7,37:39])),
  caption = '`POBTOT` table with blocks 1 to 7, and `variable table`  with a categorical variable (P_18A24) showing `missing population `NA` cells from the same blocks.',
  row.names = TRUE
  )
```

### The `block` conditional

The totals by block and gender allows to specify a rule to make sure that the sum of each categorical variable by block does not overestimate the population in the previews conditionals (`while` and `if` conditions). For that reason, a `smrow_` function is used to sum the rows that belongs to the targeted categorical variables which sum matches the total population in the block. The rules are the following:

   Gender   | Function
------------- | -------------
Male+female  | `(POBTOT[i] > smrow_mf(i))`
Male  | `(POBMAS[i] > smrow_m(i))`
Female  | `(POBFEM[i] > smrow_f(i))`

   Where   | Description
------------- | -------------
`POBTOT, POBMAS, POBFEM`  | Total population, total males and females.
`smrow_`  | Function that sums all rows of the categorical variables at the iteration cell based on gender.

Once the `smrow_` function  provides the outcome of the rows' sum in the iteration of the cell, the value is compared with the respective total in the block, if the total is bigger than the row sums, a sum of `vec.n[i] <- vec.n[i]+1` is allowed.

Note that the sum of all rows with the header format of `P_#A#` (e.g. `P_18A24`) is equal to the POBTOT of that block. The same corresponds to male and female functions.

## 2.3 Extra conditionals (M, F)

There are some extra conditions that apply specifically for the male column and others for the female column which are explained next.

### Male conditionals

The male column is in the middle between totals and females, which follows the order of the execution of the function. Some times, female values were already provided, and the totals columns were defined. The before implies that male `NA` blocks where female are known, the male cells must be calculated before a regular iteration `while` conditional. To solve that issue, the following lines of codes are provided:

```{r highlight=TRUE, eval=FALSE}
!is.na(vec.f[i])#For female precondition for male cells
#First fill all values where total and females are defined
& (((vec.mf + vec.mf.n)[i] - vec.f[i]) != vec.m.n[i])
```

### Female conditionals

In the case of the female column, sometimes the blocks of totals and male already filled the space of the block, which means there is no more space for female. The conditional verifies which is the status of the totals and males to select priority blocks where female are required. The following code solves that issue.

```{r highlight=TRUE, eval=FALSE}
#Special condition: verify that male[i] keeps under total MF[i]
& ( (vec.mf+vec.mf.n)[i] != ((vec.m+vec.m.n)[i] + vec.f.n[i]))
#Special condition: Calculate difference of new totals
& (( (vec.mf+vec.mf.n)[i] - (vec.m+vec.m.n)[i]) != 0)
```


# 3 Generic cleaning function

Finally, the declaration of the generic function that was built to clean the data for all the columns of the categorical variables `Age-ranges and sex` is declared as follows:

```{r highlight=TRUE, eval=FALSE}
clean.pvar <- function( = hpcons$P_12YMAS,
                       vec.m = hpcons$P_12YMAS_M,
                       vec.f = hpcons$P_12YMAS_F,
                       t.tot.mf = hptotcons$P_12YMAS,
                       t.tot.m = hptotcons$P_12YMAS_M,
                       t.tot.f = hptotcons$P_12YMAS_F,
                       nam.mf.n = "P_12YMAS.f",
                       nam.m.n ="P_12YMAS_M.f",
                       nam.f.n ="P_12YMAS_F.f")
```

As mentioned in section 2.1, each categorical variable has an specific generic function. 

## 3.1 Results

In the process of data transformation, it was possible to verify that negative values are not present from the differences of categorical values. It was also verified that the sum of all the values matches the totals of columns, and rows. 

Here is a summary before data cleaning for the variable `P_18A24_F`:
```{r echo=FALSE}
summary(hpcons.orig$P_18A24_F)
```
Summary after cleaning `P_18A24_F` (eliminated `NA` values):
```{r echo=FALSE}
summary(hpcons1$fP_18A24)
ggplot(hpcons1, aes(x = fP_18A24)) + geom_bar(fill = "pink")
```

Here is a summary before data cleaning for the variable `P_18A24_M`:
```{r echo=FALSE}
summary(hpcons.orig$P_18A24_M)
```

Summary after cleaning `P_18A24_M` (eliminated `NA` values):
```{r echo=FALSE}
summary(hpcons1$mP_18A24)
ggplot(hpcons1, aes(x = mP_18A24)) + geom_bar(fill = "blue")
```

```{r echo=FALSE}
hpcons0[,1:9]
```

